### 2017-07-06 paper/app.tex

The individual steps in the above computation are justified as follows: in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level in step $k$ is $\fixn{j}$ it holds that $\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s whose level in step $k$ is $>\fixn{j}$ it holds $\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$. Moreover, for all $\omega\in D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow U(\omega)\geq k+1$. Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, but there are $k-\fixn{i}$ such steps, a contradiction. This concludes the proof of~\eqref{eq:lexrsm-soundness-main}.

>>>>

The individual steps in the above computation are justified as follows: 
in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level 
in step $k$ is $\fixn{j}$ it holds that 
$\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s 
whose level in step $k$ is $>\fixn{j}$ it holds 
$\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$. Moreover, for all $\omega\in 
D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow 
U(\omega)\geq k+1$. Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that 
all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, 
since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at 
least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, 
but there are $k-\fixn{i}$ such steps, a contradiction. This concludes the 
proof of~\eqref{eq:lexrsm-soundness-main}.

\section{Complexity Clarification for Theorem~\ref{THM:ALGO}}

Since instances of linear programming is in~\textsc{PTIME}, it remains to show 
that each system $\linsystem_{\tilde{\tau}}$ is constructible in polynomial 
time. In~\cite{CFNH16:prob-termination} it shown that this can be done provided 
that guard of each transition in pCFG is a propositionally linear predicate. 
Now all transition guards in $\pCFG_\program$ are of the form $\phi$ or 
$\neg\phi$, where $\phi$ is a guard of a conditional or of while-loop in 
$\program$. If $\phi$ is a linear assertion, then $\neg\phi$ can be converted 
into a propositionally linear predicate in polynomial time, after which the 
construction of~\cite{CFNH16:prob-termination} can be used.

\section{Additional Computation for Example~\ref{EX:UNIFORM}}

 To see that $x$ is uniformly integrable in the left program and not in the 
 right one (within the inner loop), 
 imagine the inner loop as a stand-alone program and let $X_n$ be the value of 
 variable $x$ after $n$ steps of this stand-alone program (i.e., when the loop 
 terminates $x$ no longer changes). Solving a simple linear recurrence shows 
 that in the right program $\E[X_n] \rightarrow 0$ as $n\rightarrow \infty$, 
 which in particular shows uniform integrability of $X_0,X_1,X_2,\dots$. On the 
 other hand, in the left program for each $K>0$ we have $\probm(X_K \geq 
 2^K\cdot{x_0})\geq \frac{1}{2^K}$, where $x_0$ is the value of $x$ upon 
 entering the loop. Hence $\E[|X_K|cdot\indicator{X_K 
 	\geq K}]\geq 1$ for each $K$ sufficiently large, which is incompatible with 
 uniform integrability.

### 2017-07-06 paper/conclu.tex
small changes

\smallskip\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work.

>>>>

\smallskip\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied 
in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied 
in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17,MM16:proofrule-arxiv}. 
Supermartingales were also 
considered for other liveness and safety 
properties~\cite{CVS16:martingale-recurrence-persistence,BEFH16:doob}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work.
